---
title: "Comparison of exaggeration across methods"
description: "In this document, I investigate whether we expect to see more exaggeration for some of the causal identification strategies than for the others."
output: 
  distill::distill_article:
    toc: true
editor_options: 
  chunk_output_type: console
---

<style>
body {
text-align: justify}
</style>

```{r setup_comparison, include=FALSE, results='hide', warning=FALSE}
library(knitr)
opts_chunk$set(fig.path = "images/comparison/",
               cache.path = "cache/",
               cache = FALSE,
               echo = TRUE, #set to false to hide code
               message = FALSE,
               warning = FALSE,
               out.width = "85%",
               dpi = 300,
               fig.align = "center",
               dev.args = list(bg="transparent"))  
```  

```{r packages_comparison, message=FALSE, warning=FALSE, code_folding='Show packages used in this document'}
library(tidyverse) 
library(knitr) 
library(mediocrethemes)
library(here)
library(retrodesign)
library(haven)

set_mediocre_all(pal = "coty")
```

There are two ingredients in the recipe for exaggeration:

1. Publication bias
1. Low statistical power

A low statistical power leads the estimator to be imprecise and statistitcally significant estimates to be located in the tail of the distribution of estimates and to thus overestimate true effects. Publication bias leads to a larger probability of publishing estimates that come from this non-representative sample of estimates. In the absence of publication bias, low statistical power would not create exaggeration.

In order to explore whether some causal identification methods are more prone to exaggeration than others, we can thus explore which ones are more subject to each of these ingredients. 

## Differences of publication bias across methods

For publication bias, the question has been investigated by [Brodeur et al. (2020)](https://www.aeaweb.org/articles?id=10.1257/aer.20190687). By investigating the universe of articles published in 25 top economics journals, they find that IV and to a lesser extent DiD are particularly problematic in this regards. RDD and RCT seem to be less prone to this issue.

## Differences of statistical power across methods

Then, I investigate whether papers using some methods display lower statistical power than others. 

### Based on Brodeur et al. (2020)

To do so, I first run a naive analysis, using Brodeur et al.'s data. I **explore whether the design of each study would enable to retrieve the true effect if it was indeed twice smaller that the one found in the study**.

Note that [Gelman and Carlin (2014)](https://journals.sagepub.com/doi/10.1177/1745691614551642), the foundational article on exaggeration, warns against using estimate from the study as true effects as they might be exaggerated. It instead recommends using estimate from other studies or meta analyses. Since meta-analyses estimates are not readily available for the studies in Brodeur et al. (2020), we however run an analysis where hypothetical true effects are based on the published estimates. This might still be informative if we see important disparities across methods.

To run the analysis, I first retrieve the Brodeur et al. (2020) data from [OPENIPCSR](https://www.openicpsr.org/openicpsr/project/120246/version/V2/view) and store it locally. I load it and wrangle it. The authors fixed some of their data after the publication of the article (changes explained in the `Changelog.txt` file). I keep the most recent data.

```{r wrangle_brodeur}
data_brodeur_raw <- read_dta(here("inputs", "brodeur_v2", "data", "MM Data.dta"))

data_brodeur <- data_brodeur_raw |> 
  filter(pv <= 0.05) |>
  mutate(
    estimate = ifelse(!is.na(mu_new), mu_new, mu_orig),
    estimate = abs(estimate), #sign does not matter here
    se = ifelse(!is.na(sd_new), sd_new, sd_orig)
  ) |> 
  select(method, estimate, se, pv) |> 
  filter(se > 0)
```

Then, using the `retrodesign` package, I compute the statistical power and exaggeration if the true effect was in fact equal to half of the published estimate.

```{r power_brodeur}
power_brodeur <- data_brodeur |> 
  mutate(
    retro = map2(estimate/2, se, \(x, y) retro_design(x, y)), 
    #retro_design returns a list with power, typeS, typeM
  ) |> 
  unnest_wider(retro) |> 
  mutate(power = power * 100) |> 
  rename(exagg = typeM) |> 
  select(-typeS)
```

I then compute the mean power and exaggeration for each identification strategy.

```{r mean_power_brodeur, code_folding='Show the code used to generate the table'}
power_brodeur |> 
  group_by(method) |> 
  summarise(
    mean_power = mean(power),
    mean_exagg = mean(exagg)
  ) |> 
  rename_with(\(x) str_to_title(str_replace_all(x, "_", " "))) |> 
  kable()
```

There does not seem to be substantial differences in means. If anything, RCT seem to perform slightly less well than the others. 

I then plot the corresponding distributions of power and exaggeration, split by method

```{r graph_power_brodeur, code_folding='Show the code used to generate the graph'}
power_brodeur |> 
  ggplot(aes(x = power)) +
  geom_histogram(bins = 40) +
  facet_wrap(~method, scales = "free_y") + 
  labs(
    title = "Distribution of hypothetical statistical power",
    subtitle = "Comparison across identification strategies",
    x = "Power (%)",
    y = "Count",
    caption = "Hypothetical true effect equals to half of the estimate"
  ) 
  # theme(legend.position = "none")

power_brodeur |> 
  ggplot(aes(x = power, fill = method, color = method)) +
  geom_density(alpha = 0.1, adjust = 0.5) +
  # facet_wrap(~method, scales = "free_y") + 
  labs(
    title = "Distribution of hypothetical statistical power",
    subtitle = "Comparison across identification strategies",
    x = "Power (%)",
    y = "Density", 
    caption = "Hypothetical true effect equals to half of the estimate",
    fill = NULL,
    color = NULL
  )

power_brodeur |> 
  filter(exagg < 3) |> 
  ggplot(aes(x = exagg, fill = method, color = method)) +
  geom_density(alpha = 0.1) +
  # facet_wrap(~method) + 
  labs(
    title = "Distribution of hypothetical exaggeration",
    subtitle = "Comparison across identification strategies",
    x = "Exaggeration ratio",
    y = "Density", 
    caption = "Hypothetical true effect equals to half of the estimate
    Estimates with an exaggeration ratio greater than 3 are filtered out",
    fill = NULL,
    color = NULL
  )
```

The results are rather similar across strategies. IF anything, we notice that IV and RCT perform slightly less well than DID and RDD.

However, if there is initially more exaggeration for one of the methods, considering hypothetical true effect sizes equal to half of the observed estimates will be dismissive since we would not be comparing similar things.








