---
title: "Intuition, experimental studies and replication crisis"
description: "In this document, we analyze replications of experimental studies to illustrate the consequences of low statistical power in economics and give intuition for the trade-off we underline."
author:
  - name: Vincent Bagilet 
    url: https://vincentbagilet.github.io/
    affiliation: Columbia University
    affiliation_url: https://www.columbia.edu/
  - name: Léo Zabrocki 
    url: https://www.parisschoolofeconomics.eu/en/
    affiliation: Paris School of Economics
    affiliation_url: https://www.parisschoolofeconomics.eu/en/
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
editor_options: 
  chunk_output_type: console
---

<style>
body {
text-align: justify}
</style>

```{r setup_experiments, include=FALSE, results='hide', warning=FALSE}
library(knitr)
opts_chunk$set(fig.path = "images/",
               cache.path = "cache/",
               cache = FALSE,
               echo = TRUE, #set to false to hide code
               message = FALSE,
               warning = FALSE,
               out.width = "85%",
               dpi = 300,
               fig.align = "center",
               dev.args = list(bg="transparent"))  
```  

```{r packages_experiments, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse) 
library(knitr) 
library(mediocrethemes)
library(haven)
library(here)
library(readxl)
library(retrodesign)
# library(gganimate)

set.seed(3)

set_mediocre_all(pal = "coty")
```


## Intuition for exaggeration

To illustrate the consequences of low power, we analyze the results from [Camerer et al (2016)](https://www.science.org/doi/10.1126/science.aaf0918) through the prism of statistical power. The authors of this paper replicated laboratory experiments in economics.

They report their replication results, alongside the original results, on the [project website](http://experimentaleconreplications.com/scripts.html), we only need to run their Stata script `create_studydetails.do` to recover all the information. 

We then compute the power of the initial analysis if the true effect is in fact equal to the replication's.

```{r camerer_retro, echo=FALSE}
rep_camerer <- read_dta("Inputs/studydetails.dta") %>% 
  mutate(
    seorig = (eorigu95 - eorig)/qnorm(0.975),
    serep = (erepu95 - erep)/qnorm(0.975)
  )

retro_camerer <- rep_camerer %>% 
  select(A = erep, s = seorig) %>% 
  # select(A, s) %>% 
  pmap_dfr(retrodesign) %>% 
  cbind(rep_camerer) %>% 
  as_tibble()
```

### Analyzing all studies

We quickly plot and analyze the results obtained, *ie* the distribution of the exaggeration ratio and power.

```{r camerer_retro_analysis, echo=FALSE}
retro_camerer %>% 
  ggplot() +
  geom_dotplot(aes(x = exaggeration), color = NA) +
  labs(
    title = "Distribution of the exaggeration ratio in the original studies",
    subtitle = "If the true effect was equal to the replicated one",
    x = "Exaggeration ratio (log scale)",
    y = "Number of studies",
    caption = "Each dot represents one study"
  ) + 
  scale_x_log10() +
  scale_y_continuous(NULL, breaks = NULL) 

retro_camerer %>%
  ggplot() +
  geom_dotplot(aes(x = power*100), color = NA) +
  labs(
    title = "Distribution of the power in the original studies",
    subtitle = "If the true effect was equal to the replicated one",
    x = "Power (%)",
    y = "Number of studies",
    caption = "Each dot represents one study"
  ) +
  scale_y_continuous(NULL, breaks = NULL) 
# 
# retro_camerer %>% 
#   ggplot() +
#   geom_histogram(aes(x = power))
# 
# retro_camerer %>% 
#   count(exaggeration > 1.5)
```

The median power would be `r round(median(retro_camerer$power),2)`%. The median replicated estimates is equal to  `r round(median(1/retro_camerer$exaggeration), 2)` times the original estimate.

We then compute the number and proportion of original studies that were statistically significant. 

```{r echo=FALSE}
rep_camerer %>% 
  mutate(
    original_significant = ifelse(porig < 0.05, "Yes", "No")
  ) %>% 
  group_by(original_significant) %>% 
  summarise(
    nb = n(),
    prop = n()/nrow(.)
  ) %>% 
  kable(col.names = c(
    "Original estimate statistically significant", 
    "Number",
    "Proportion"),
    digits = 2
  )
```

We then do the same thing for the replication studies.

```{r echo=FALSE}
rep_camerer %>% 
  mutate(
    rep_significant = ifelse(prep < 0.05, "Yes", "No")
  ) %>% 
  group_by(rep_significant) %>% 
  summarise(
    nb = n(),
    prop = n()/nrow(.)
  ) %>% 
  kable(col.names = c(
    "Replication estimate statistically significant", 
    "Number",
    "Proportion"),
    digits = 2
  )
```

We then compute the proportion of original studies that would have adequate power as defined by the customary and arbitrary 80% threshold, still assuming that the true effect is equal to the replication one.

```{r echo=FALSE}
retro_camerer %>% 
  mutate(
    adequate_power = ifelse(power >= 0.8, "Yes", "No")
  ) %>% 
  group_by(adequate_power) %>% 
  summarise(
    nb = n(),
    prop = n()/nrow(.)
  ) %>% 
  kable(col.names = c(
    "Adequate power", 
    "Number",
    "Proportion"),
    digits = 2
  )
```

### Focus on one particular study

Here, we focus on one particular study in order to illustrate in more details the problem at play. We want to simulate what could have yielded replication of the initial study if the true effect was equal to the replication estimate.

We select one of the studies, Abeler et al. (2011), that we initially selected at random and draw the graph of interest. The way we calculated the standard error is not perfectly accurate so we use the information available in the [replication report](http://experimentaleconreplications.com/replicationreports.html).

```{r graph_retrodesign_camerer, echo=FALSE, fig.height=4.3, fig.width=9, out.width=1200, dpi = 700}
random_study <- rep_camerer %>% 
  slice(1) 

n_iter <- 500

data_graph_distrib <- rnorm(random_study$erep, random_study$seorig, n = n_iter) %>% 
  as_tibble() %>% 
  mutate(
    n = row_number(),
    non_significant = dplyr::between(
      value, 
      - 1.96*sd(value), 
      1.96*sd(value)
    ),
    significant = ifelse(non_significant, "Non significant", "Significant") 
  ) 

graph_original <- data_graph_distrib %>% 
  ggplot(aes(x = n, y = value, color = significant)) + 
  #original study
  geom_point(aes(x = -30, y = random_study$eorig), color = "darkred", size = 2) +
  geom_linerange(aes(
    x = -30,
    ymin = random_study$eorig - 1.96*random_study$seorig,
    ymax = random_study$eorig + 1.96*random_study$seorig), color = "darkred") +
  labs(
    title = "Illustration of the exaggeration and power issues",
    subtitle = "500 draws of an estimate ~ N(Effect size in replication, std err in original study)",
    x = "Draw",
    y = "Point estimate"
  ) +
  scale_color_discrete(name = "") +
  xlim(c(-30, 500)) +
  ylim(c(-0.25, 0.4))

ggsave("graph_retrodesign_camerer_1.pdf", path = "images", width = 9, height = 4.5)

graph_rep <- graph_original +
  #replication 
  geom_point(aes(x = -20, y = random_study$erep), color = "darkblue", size = 2) +
  geom_linerange(aes(
    x = -20,
    ymin = random_study$erep - 1.96*random_study$serep,
    ymax = random_study$erep + 1.96*random_study$serep), color = "darkblue") 

ggsave("graph_retrodesign_camerer_2.pdf", path = "images", width = 9, height = 4.5)

graph_rep_original <- graph_rep +
  #repliccation with design of the original study
  geom_point(aes(x = -10, y = random_study$erep), color = "gray50", size = 2) +
  geom_linerange(aes(
    x = -10,
    ymin = random_study$erep - 1.96*random_study$seorig,
    ymax = random_study$erep + 1.96*random_study$seorig), color = "gray50")

ggsave("graph_retrodesign_camerer_3.pdf", path = "images", width = 9, height = 4.5)

graph_rep_original <- graph_rep_original +
  # geom_hline(aes(yintercept = mean(value)), size = 0.8) +
  geom_point(alpha = 0.8) +
  geom_hline(aes(yintercept = 0), size = 0.3, linetype = "solid") 
  # transition_layers(layer_length = 2, from_blank = FALSE) #to animate

ggsave("graph_retrodesign_camerer_4.pdf", path = "images", width = 9, height = 4.5)

graph_rep_original
```

- In red is the estimate from the original study and its 95% confidence interval
- The estimate is significant and has been published. Yet, it is pretty noisy.
- In blue is the estimate from the replicated study and its 95% confidence interval
- We notice that this second estimate is both more precise and smaller than the initial one. It still remains noisy
- Let's assume that the true effect is actually equal to this second estimate (note that this is unlikely)
- Would the design of the initial study be good enough to detect this true effect? *ie* if we replicated the initial study, could we reject the null of no effect (knowing that the true effect is equal to the replicated estimate)
- In gray is the estimate form the replicated study but with a standard error equal to the initial study's (approximately the standard errors that would have been obtained with the design of the initial study)
- This estimate is non significant. In this instance, we would not have been able to reject the null of no effect
- Now, if we replicate this study `r n_iter` times, running `r n_iter` lab experiments, in some cases we would get statisitcally significant estimates (the beige dots) and in some others non statistically significant ones (the green dots)
- If we would have been a bit more lucky, we could have gotten a sample of individuals that would have yielded one of the beige estimates
- Now, we notice that, on average, statistically significant estimates overestimate the true effect by a factor `r data_graph_distrib %>% filter(!non_significant) %>% .$value %>% mean() / random_study$erep` (average of `r data_graph_distrib %>% filter(!non_significant) %>% .$value %>% mean()` while the true effect is `r random_study$erep`). Gelman and Carlin call this inflation factor type M error.
- In this case, the power is basically the proportion of statistically significant estimates
- If the study had more power, the sd would be smaller and most estimates would be statistically significative (because there is indeed a non null effect)
- But since the power is low, if by chance the sample of individuals we get yields a statistically significant estimate, this estimate will overestimate the true effect


<!-- ## Randomized Control Trials -->

<!-- We want to look at replications of RCTs in Development Economics. To do so, we use [the list of replication papers put together by Sandip Sukhtankar](https://www.aeaweb.org/articles?id=10.1257/aer.p20171120).  -->

<!-- We gather the list of RCTs that have been replicated in Development Economics. -->

<!-- ```{r rep_dvpt} -->
<!-- rep_dvpt <- read_dta(here("Misc", "replication_data_final.dta")) -->

<!-- # rep_dvpt %>%  -->
<!-- #   filter((RCT == "Yes") & (Replicated == "Replicated")) %>%  -->
<!-- #   .$ReplicationPaperTitle -->
<!-- ``` -->

## Intuition for the impact of precision on exaggeration

The trade-off we describe between exaggeration and confounders arises due to differences in precision between estimators. To illustrate this, we generate estimates from two unbiased estimators with identical mean but different variances and plot their distributions, distinguishing between estimates that are and are not significant.

```{r intuition_trade_off, echo=FALSE, fig.asp=1.2}
N <- 10000
sd_precise <- 0.23
sd_imprecise <- 0.7
true_effect <- 1

data_intuition <- 
  tibble(
    estimate = rnorm(N, true_effect, sd_precise), 
    precise = "Precise estimator"
  ) %>% 
  rbind(
    tibble(
      estimate = rnorm(N, true_effect, sd_imprecise), 
      precise = "Imprecise estimator"
    )
  ) %>% 
  group_by(precise) %>% 
  mutate(sd = sd(estimate)) %>% 
  ungroup() %>% 
  mutate(
    signif = abs(estimate) > sd*1.96,
    signif_name = ifelse(signif, "Significant", "Non significant"),
    signif_estimate = ifelse(signif, estimate, NA)
  ) %>% 
  group_by(precise) %>% 
  mutate(
    estimate_sign = ifelse(signif, estimate, NA),
    mean_signif = mean(estimate_sign, na.rm = TRUE)
  ) %>% 
  ungroup()

data_intuition %>% 
  ggplot(aes(x = estimate, fill = signif_name)) +
  geom_histogram(bins = 100) +
  # geom_area(stat = "bin", bins = 40) +
  facet_wrap(~ precise, nrow = 2) +
  geom_vline(xintercept = 0, linetype = "solid", linewidth = 0.1) +
  geom_vline(xintercept = true_effect) +
  geom_vline(
    aes(xintercept = mean_signif), 
    color = "#976B21", 
    linetype = "solid", 
    size = 0.3
  ) + 
  labs(
    x = "Estimate",
    y = "Count",
    title = "Exaggeration of an imprecise estimator",
    subtitle = "Distribution of two estimators with same mean and different variance",
    fill = NULL,
    caption = "The green dashed line represents the true effect size
      The brown solid line represents the average of significant estimates"
  )

ggsave(here("images", "graph_intuition_precision.pdf"), width = 6, height = 5)
```

In the case of the precise estimator, the mean of statistically significant estimates is almost equal to the true effect. The estimator is unbiased and almost all estimates are significant. However, in the case of the imprecise estimator, only a sub-sample of the estimates are significant: those that are far away enough from 0 (at least 1.96 s.d.). As a consequence, the mean of  statistically significant estimates is different from the true effect, even if the estimator is unbiased. This hints to the trade-off between biased estimates from naive regressions and less precise causal inference estimates.



