---
title: "Mathematical derivation"
description: "In this document, we summarise the math derivation described in appendix of the paper and run simple simulations to both illustrate and check our closed form solution."
author:
  - name: Vincent Bagilet 
    url: https://vincentbagilet.github.io/
    affiliation: Columbia University
    affiliation_url: https://www.columbia.edu/
  - name: Léo Zabrocki 
    url: https://www.parisschoolofeconomics.eu/en/
    affiliation: Paris School of Economics
    affiliation_url: https://www.parisschoolofeconomics.eu/en/
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
editor_options: 
  chunk_output_type: console
---

<style>
body {
text-align: justify}
</style>

```{r setup_RDD, include=FALSE, results='hide', warning=FALSE}
library(knitr)
opts_chunk$set(fig.path = "images/",
               cache.path = "cache/",
               cache = FALSE,
               echo = TRUE, #set to false to hide code
               message = FALSE,
               warning = FALSE,
               out.width = "85%",
               dpi = 300,
               fig.align = "center",
               dev.args = list(bg="transparent"))  
```  

```{r packages_maths, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(mediocrethemes)
library(broom)
library(ggdag)
library(AER)
library(tictoc)
library(beepr)
library(here)

set_mediocre_all(pal = "coty")
```

The full proof is available here.



## Simulations

### Modeling choices

We then run simulations to both illustrate and check these results. The structure of the code is similar to our other simulations: we first set baseline parameters, define the set of parameters to consider and then run the simulations. A simpler code could be written (with a simple for loop) but we chose this coding practice for consistency.

The DGP is described in the following DAG:

```{r DAG_maths, echo=FALSE, fig.asp=0.25, out.width='70%'}
dagify(Y ~ X + W,
       X ~ W + Z,
       outcome = "Y",
       latent = "W",
       coords = list(x = c(Y = 3, X = 2, W = 2.5, Z = 1),
                     y = c(Y = 1, X = 1, W = 0, Z = 1))) %>% 
  # ggdag() +
  ggplot(ggplot2::aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(alpha = 0, stroke = TRUE) +
  geom_dag_edges() +
  geom_dag_text(
    colour = "black", 
    size = 7,
    family = "Lora"
  ) +
  theme_void()
```

As in the math section we have:

$$y_{i} = \beta_{0} + \beta_1 x_{i} + \delta w_{i} + u_{i}$$
$$x_{i} = \mu_x + \eta z_{i} + \gamma w_{i} + \epsilon_{i}$$

### Generate the data

As in the method specific simulations, we create three function, one to generate the data, one to run the estimation and one to compute a simulation (just bundling together the two previous functions).

For simplicity, we consider that $w$ and $z$ is normally distributed. We also consider normally distributed error terms. As a linear combination of normally distributed variables, $x$ and $y$ will also be normally distributed.

```{r generate_data_maths}
generate_data_maths <- function(N,
                                mu_z,
                                sigma_z,
                                sigma_w,
                                sigma_epsilon,
                                alpha,
                                eta,
                                gamma,
                                sigma_u,
                                beta_0,
                                beta_1,
                                delta) {
  
  data <- tibble(id = 1:N) %>%
    mutate(
      # D = rbernoulli(N, ptreat),
      z = rnorm(N, mu_z, sigma_z),
      w = rnorm(N, 0, sigma_w),
      epsilon = rnorm(N, 0, sigma_epsilon),
      x = alpha + eta*z + gamma*w +  epsilon,
      u = rnorm(N, 0, sigma_u),
      y = beta_0 + beta_1*x + delta*w + u
    )
}
```

We then define the baseline parameters values. The numbers chosen are not the usual round ones. We build the present simulations to check our mathematical derivations and non-round numbers fulfill this role better than round numbers. Yet, the magnitudes are chosen in order to mimic a study with low statistical power. 

```{r param_maths, echo=FALSE}
# baseline_param_maths <- tibble(
#   N = 500,
#   mu_x = 0.3,
#   beta_0 = 0.2,
#   beta_1 = 0.1,
#   delta = 0.5,
#   gamma = 0.5,
#   eta = 1,
#   sigma_epsilon = 1,
#   sigma_u = 1,
#   sigma_w = 1,
#   sigma_z = 0.5
# )

baseline_param_maths <- tibble(
  N = 1000,
  mu_z = 0, 
  sigma_z = 0.72,
  sigma_w = 0.6,
  sigma_epsilon = 1.2,
  mu_x = 0.4,
  eta = 1.2,
  gamma = 0.2,
  sigma_u = 0.8,
  beta_0 = 0.2,
  beta_1 = 0.1,
  delta = 0.3
)

baseline_param_maths %>% kable()

n_iter <- 1000

param_maths <- baseline_param_maths %>% 
  crossing(rep_id = 1:n_iter) %>% 
  select(-rep_id)

test_data <- pmap_dfr(baseline_param_maths, generate_data_maths)
```

We then run the estimations. We consider four separate models:

- Unbiased: the true model, regressing $y$ on $x$ and $w$, therefore without omitted variable
- Biased: the "naive" regression of $y$ on $x$, therefore omitting for $w$
- IV: the 2SLS regression of $y$ on $x$ instrumented by $z$
- Reduced form: regressing $y$ on $d$

We also write a short function to wrangle the regression output to a convenient format, `lint_reg_maths`.

```{r estimate_maths}
lint_reg_maths <- function(reg_out) {
  name_model <- deparse(substitute(reg_out)) %>% 
    str_extract("(?<=reg_).+") 
    
  reg_out %>%
    tidy() %>%
    filter(term %in% c("x", "z")) %>%
    mutate(model = name_model) %>% 
    select(model, everything())
}

estimate_maths <- function(data) {
  #estimation
  reg_unbiased <- lm(data = data, y ~ x + w)
  reg_biased <- lm(data = data, y ~ x)
  reg_reduced_form <- lm(data = data, y ~ z)
  reg_iv <- ivreg(data = data, y ~ x | z)

  #wranglin results
  results <- lint_reg_maths(reg_unbiased) %>%
    rbind(lint_reg_maths(reg_biased)) %>%
    rbind(lint_reg_maths(reg_reduced_form)) %>%
    rbind(lint_reg_maths(reg_iv)) %>%
    mutate(signif = (p.value < 0.05))
}
```

We then bundle these two functions to compute a simulation.

```{r compute_sim_maths}
compute_sim_maths <- function(...) {
  generate_data_maths(...) %>% 
    estimate_maths()
}
```

The output of one simulation is:

```{r one_sim, echo=FALSE}
pmap_dfr(baseline_param_maths, compute_sim_maths) %>% 
  kable()
```

We then run all the simulations and save the results.

```{r run_sim_maths, eval=FALSE, echo=FALSE}
tic()
sim_maths <- pmap_dfr(param_maths, compute_sim_maths)
beep()
toc()

saveRDS(sim_maths, here("Outputs/sim_maths.RDS"))
```

### Analysis of the results

First, we plot the distribution of the estimates

```{r graph_maths, echo=FALSE, fig.asp=1.2}
sim_maths <-readRDS(here("Outputs/sim_maths.RDS"))

sim_maths %>% 
  mutate(
    signif_name = ifelse(signif, "Significant", "Non significant"),
    model_name = model %>% 
      str_to_title() %>% 
      str_replace_all("_", " ") %>% 
      str_c(., " model")
  ) %>% 
  group_by(model) %>% 
  mutate(
    estimate_sign = ifelse(signif, estimate, NA),
    mean_signif = mean(estimate_sign, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  filter(dplyr::between(estimate, -0.5, 0.7)) %>% 
  ggplot(aes(x = estimate, fill = signif_name)) +
  geom_histogram(bins = 50) +
  # geom_area(stat ="bin", bins = 50) +
  facet_wrap(~ model_name, nrow = 4) +
  geom_vline(xintercept = 0, linetype = "solid", size = 0.1) +
  geom_vline(xintercept = baseline_param_maths$beta_1) +
  geom_vline(
    aes(xintercept = mean_signif), 
    color = "#976B21", 
    linetype = "solid", 
    size = 0.3
  ) +
  labs(
    title = "Distribution of estimates produced in the simulation",
    subtitle = "Comparison across models",
    fill = "",
    x = "Estimate",
    y = "Count",
    caption = "The green vertical dashed line represents the true effect size
      The brown vertical solid line represents the average of significant estimates"
  ) 
```

We then look at the mean of the estimates and of the statistically significant ones for each model:

```{r mean_estimates_maths, echo=FALSE}
sim_maths %>% 
  mutate(est_sign = ifelse(signif, estimate, NA)) %>% 
  group_by(model) %>% 
  summarise(
    mean_sign = mean(est_sign, na.rm = TRUE),
    mean_all = mean(estimate)
  ) %>%  
  mutate(true_effect = baseline_param_maths$beta_1) %>% 
  kable(col.names = str_to_title(str_replace(names(.), "_", " ")))
```

Visible both on the graph and the table, we get that:
- The unbiased model is indeed unbiased as the estimates are centered on the true effect
- The IV and the reduced form are unbiased, *ie* centered on the true effect. However, statistically significant estimates are of average exaggerating the true effect size. In if there is publication bias, published estimates would in this case overestimate the true effect.
- The biased model is indeed biased but more precise than the IV and reduced form

Of course these results depend on the choice of parameters but this simulation has an illustrative purpose only.

## Checking the maths

### Biased model

Using the simulations, we can check the maths. Based on our derivations, we expect to have:

$$ \hat{\beta}_{ovb} \overset{d}{\to} \mathcal{N}\left( \beta_1 + \delta \gamma \dfrac{\sigma_{w}^{2}}{\sigma_{x}^{2}}, \ \dfrac{\sigma_{v}^{2}}{n \cdot \sigma_{x}^{2}} \right) $$
First, we need to compute the variance of $v$ and $x$. We verify that our formulas are correct: 

- We expect to have $\sigma_x^2 = \eta^{2}\sigma_{z}^{2} + \gamma^2\sigma_w^2 + \sigma_{\epsilon}^2$ since $\text{Var}(d) = p_T(1 - p_T)$. For our parameters values, it corresponds to `r  baseline_param_maths %>% mutate(th_var_x = eta^2*sigma_z^2 + gamma^2*sigma_w^2 + sigma_epsilon^2) %>% .$th_var_x`. In an example dataset with 20000 observations, the observed variance of $x$ is `r pmap_dfr(baseline_param_maths %>% mutate(N = 2000000), generate_data_maths) %>% .$x %>% var()`. The simple formula for the variance of $x$ seems to be correct.
- We expect to have $\sigma_v^2 = \delta^2\sigma_w^2 + \sigma_u^2$. For our parameters values, it corresponds to 
`r  baseline_param_maths %>% mutate(th_var_v = delta^2*sigma_w^2 + sigma_u^2) %>% .$th_var_v`. In an example dataset with 20000 observations, the observed variance of the residuals of the regression of $y$ on $x$ is `r pmap_dfr(baseline_param_maths %>% mutate(N = 20000), generate_data_maths) %>% lm(data = ., y ~ x) %>% .$residuals %>% var`. The simple formula for the variance of $v$ seems to be correct.

Then, we can compute the theoretical variance for the mean and the variance of the estimator and compare them to the ones observed in the simulations

```{r check_moments_biased, echo=FALSE}
th_moments_biased <- baseline_param_maths %>% 
  mutate(
    th_sigma_x = sqrt(eta^2*sigma_z^2 + gamma^2*sigma_w^2 + sigma_epsilon^2),
    th_sigma_v = sqrt(delta^2*sigma_w^2 + sigma_u^2),
    mean = beta_1 + delta*gamma*sigma_w^2/th_sigma_x^2,
    variance = th_sigma_v^2/(N*th_sigma_x^2)
  ) %>% 
  select(mean, variance) %>% 
  mutate(type = "Theoretical")

sim_moments_biased <- sim_maths %>% 
  filter(model == "biased") %>%
  # group_by(model) %>%
  summarise(
    mean = mean(estimate),
    variance = var(estimate)
  ) %>% 
  mutate(type = "Simulated")

rbind(th_moments_biased, sim_moments_biased) %>% 
  select(type, mean, variance) %>%
  mutate(model = "Biased") %>% 
  kable(col.names = str_to_title(names(.)))
```

### IV

```{r check_moments_iv, echo=FALSE}
th_moments_iv <- baseline_param_maths %>% 
  mutate(
    mean = beta_1,
    variance = (delta^2*sigma_w^2 + sigma_u^2)*((1-eta)*mu_z^2 - alpha*mu_z + sigma_z^2)/(N*eta^2*sigma_z^4)
  ) %>% 
  select(mean, variance) %>% 
  mutate(type = "Theoretical")

sim_moments_iv <- sim_maths %>% 
  filter(model == "iv") %>%
  # group_by(model) %>%
  summarise(
    mean = mean(estimate),
    variance = var(estimate)
  ) %>% 
  mutate(type = "Simulated")

rbind(th_moments_iv, sim_moments_iv) %>% 
  select(type, mean, variance) %>% 
  mutate(model = "IV") %>% 
  kable(col.names = str_to_title(names(.)))
```



```{r var_estimates_maths, echo=FALSE}
var_est <- sim_maths %>% 
  group_by(model) %>% 
  summarise(var_estimate = var(estimate))

var_est %>% 
  kable(col.names = str_to_title(str_replace(names(.), "_", " ")))

var_est[[which(var_est$model == "iv") , "var_estimate"]]/var_est[[which(var_est$model == "biased") , "var_estimate"]]

baseline_param_maths %>% 
  mutate(
    ratio_var = 1 + (gamma^2*sigma_w^2+sigma_epsilon^2)/(eta^2*sigma_z^2),
    var_beta_iv = (delta^2*(sigma_w^2) + sigma_u^2)*((0)^2+1/(N*eta^2*sigma_z^2))
  ) %>% 
  select(var_beta_iv, ratio_var)
```



