---
title: "Visualization tool"
description: "In this document, I build an example to describe the implementation and usage of the visualization tool outlined in this paper."
output: 
  distill::distill_article:
    toc: true
editor_options: 
  chunk_output_type: console
---
ı
<style>
body {
text-align: justify}
</style>

```{r setup_viz, echo=FALSE, results='hide', warning=FALSE}
library(knitr)
opts_chunk$set(fig.path = "images/viz/",
               cache.path = "cache/",
               cache = FALSE,
               echo = TRUE, #set to false to hide code
               message = FALSE,
               warning = FALSE,
               out.width = "85%",
               dpi = 300,
               fig.align = "center",
               dev.args = list(bg="transparent"))  
```  

```{r packages_viz, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse) 
library(knitr) 
library(mediocrethemes)
library(broom)
library(AER)
library(here)
library(haven)
library(fixest)

set_mediocre_all(pal = "coty")
```

## Setting

In order to present the replication tool outlined in this paper, I apply it to the case of [Deschênes and Greenstone (2007)](https://www.aeaweb.org/articles?id=10.1257/aer.97.1.354), henceforth DG. I chose this example because [Fisher et al. (2012)](https://www.aeaweb.org/articles?id=10.1257/aer.102.7.3749) and [Auffhammer et al. (2013)](https://www-journals-uchicago-edu.ezproxy.cul.columbia.edu/doi/10.1093/reep/ret016) underline that fixed effects absorb most of the variation in this particular paper.

DG analyses the impact of climate change on agricultural profits. I focus on their preferred regression, reported in column 4 of table 5 and described in their equation 4. This is described in the file `Table5_4.do` from their [replication package](https://www.openicpsr.org/openicpsr/project/116261/version/V1/view?path=/pcms/projects/1/1/6/2/116261/V1.0.1&type=project). I also got the data from this replication package. For simplicity in exposition, I abstract from any of the concerns raised in Fisher et al. (2012). These comments while valid are distinct from this document's objective of highlighting the small amount of variation used for identification.

## Data cleaning

First, I clean the data as they do in their `do` file, simply translating their `Stata` code into a language usable through an open access software, `R` (I actually asked ChatGPT to translate this for me and it did an ok job at it, I just checked the output, corrected a mistake and reformated it).

```{r clean_DG, code_folding=TRUE}
DG_data <- haven::read_dta(here("inputs", "replication_deschenes", "DATA1.dta"))

DG_clean <- DG_data |>
  mutate(
    y = (rtfsale - rtotexp) * 1000 / fland,
    dry_dd89 = ifelse(dry == 1, dd89, 0),
    dry_dd89_sq = ifelse(dry == 1, dd89 * dd89, 0),
    dry_prcp = ifelse(dry == 1, prcp, 0),
    dry_prcp_sq = ifelse(dry == 1, prcp * prcp, 0),
    irr_dd89 = ifelse(dry == 0, dd89, 0),
    irr_dd89_sq = ifelse(dry == 0, dd89 * dd89, 0),
    irr_prcp = ifelse(dry == 0, prcp, 0),
    irr_prcp_sq = ifelse(dry == 0, prcp * prcp, 0)
  )
```

## Regression

I then run one of the regressions in the paper using the function `fixest::feols()`. 

```{r}
reg <- DG_clean |> 
  fixest::feols(
    fml = y ~ regex("^dry.*") +  regex("^irr.*") + 
      regex("^x.*") + regex("^sst.*") | fips, 
    weights = ~ fland,
    se = "hetero"
        )

summary(reg) |> tidy()
```

I obtain the same results as in the paper. 

The parameter of interest of the authors is a  combination of several parameters. Since I only aim at exposing the visualization principle in this document, I pretend that the only parameter of interest is `irr_prcp`. 

## Computing the weights

To compute the weights, I compute the residuals of the regression of this variable of interest other the rest of the covariates. 

```{r viz_compute_weights}
reg_weights_fe <- feols(
  data = DG_clean, 
  fml = irr_prcp ~ regex("^dry.*") +  regex("^irr.*") + 
    regex("^x.*") + regex("^sst.*") - irr_prcp | fips + year,
  weights = ~ fland,
  se = "hetero"
)

# reg_y_fe <- feols(
#   data = DG_clean, 
#   fml = y ~ regex("^dry.*") +  regex("^irr.*") + 
#     regex("^x.*") + regex("^sst.*") - irr_prcp | fips,
#   weights = ~ fland,
#   se = "hetero"
# )

reg_weights_no <- DG_clean %>%
  feols(
  data = ., 
  fml = irr_prcp ~ regex("^dry.*") +  regex("^irr.*") - irr_prcp | year,
  weights = ~ fland,
  se = "hetero"
)

# reg_y_no <- DG_clean %>%
#   feols(
#   data = .,
#   fml = y ~ regex("^dry.*") +  regex("^irr.*") - irr_prcp,
#   weights = ~ fland,
#   se = "hetero"
# )

DG_weights <- DG_clean |>
  mutate(
    weight_fe = residuals(reg_weights_fe)^2,
    weight_no = residuals(reg_weights_no)^2,
    #normalize the weights
  ) |> 
  group_by(year) |>
  mutate(
    weight_fe = weight_fe/sum(weight_fe),
    weight_no = weight_no/sum(weight_no)
  )
```

## Visualizing the weights

```{r viz_heatmap, code_folding=TRUE, fig.asp=1.5}
DG_weights_tidy <- DG_weights %>% 
  pivot_longer(
    cols = starts_with("weight_"), 
    names_prefix = "weight_",
    names_to = "fe",
    values_to = "weight"
  ) |> 
  mutate(fe = (fe == "fe")) 

DG_weights_tidy |> 
  slice(1:2000) |>
  # filter(!fe) |> 
  mutate(fe = ifelse(fe, "With fixed effects", "Without fixed effects")) %>% 
  ggplot(aes(x = as.factor(year), y = as.factor(fips), fill = as.numeric(weight))) +
  geom_tile() +
  facet_wrap( ~ fe) + 
  # coord_fixed() + 
  labs(
    title = "Regression weights for each observation",
    subtitle = "Comparison of regressions with and without year-state fixed effects",
    y = "Individual id",
    x = NULL,
    fill = "Weight"
  ) +
  theme(
    axis.text.y = element_blank(),
    # axis.ticks.x = element_blank(),
    panel.grid.major.y = element_blank()
  ) + 
  scale_mediocre_c(pal = "coty", gradient = "left") 
```

Maps

```{r shape_files}
library(maps)


#shape file
shape_counties <- map_data("county") |> 
  as_tibble() 

#need to map to county
data(county.fips)

county_fips_tidy <- county.fips |> 
  as_tibble() |> 
  separate(polyname, into = c("region", "subregion"), sep = ",") |> 
  mutate(fips = as.numeric(fips))

shape_fips <- shape_counties |> 
  full_join(county_fips_tidy, by = c("region", "subregion")) 

```



```{r viz_maps, code_folding=TRUE, fig.asp=1}
DG_weights |> 
  group_by(year) |> 
  left_join(shape_fips, by = c("fips")) |> 
  ungroup() |> 
  ggplot() +
  geom_polygon(
    aes(x=long, y=lat, group=group, fill=weight_fe), 
    colour = "lightgray",
    size = 0.1
  ) + 
  scale_mediocre_c(pal = "coty", gradient = "left") +
  facet_wrap(~ year) +
  coord_fixed(1.3) +
  theme_mediocre_map() +
  labs(
    x = NULL, y = NULL,
    fill = "Weight",
    title = "Only a subset of the observations actually contributes to the estimation",
    subtitle = "Weights with fixed effects"
  )

DG_weights |> 
  group_by(year) |>
  left_join(shape_fips, by = c("fips")) |> 
  ungroup() |>
  ggplot() +
  geom_polygon(
    aes(x=long, y=lat, group=group, fill=weight_no), 
    colour = "lightgray",
    size = 0.1
  ) + 
  scale_mediocre_c(pal = "coty", gradient = "left") +
  facet_wrap(~ year) +
  coord_fixed(1.3) +
  theme_mediocre_map() +
  labs(x = NULL, y = NULL) +
  labs(
    x = NULL, y = NULL,
    fill = "Weight",
    title = "Only a subset of the observations actually contributes to the estimation",
    subtitle = "Weights without fixed effects"
  )
```

## New tests for maps (2023-10-24)

As a reminder, the outcome variable is y and I use `irr_prcp` as the main explanatory var of interest

```{r}
DG_clean |> 
  group_by(year) |>
  left_join(shape_fips, by = c("fips")) |> 
  ungroup() |>
  ggplot() +
  geom_polygon(
    aes(x = long, y = lat, group = group, fill = y), 
    colour = "lightgray",
    size = 0.1
  ) + 
  scale_mediocre_c(pal = "coty", gradient = "left") +
  facet_wrap(~ year) +
  coord_fixed(1.3) +
  theme_mediocre_map() +
  labs(
    title = "Raw y",
    x = NULL, 
    y = NULL
  ) 

DG_clean |> 
  group_by(year) |>
  left_join(shape_fips, by = c("fips")) |> 
  ungroup() |>
  ggplot() +
  geom_polygon(
    aes(x = long, y = lat, group = group, fill = irr_prcp), 
    colour = "lightgray",
    size = 0.1
  ) + 
  scale_mediocre_c(pal = "coty", gradient = "left") +
  facet_wrap(~ year) +
  coord_fixed(1.3) +
  theme_mediocre_map() +
  labs(
    title = "Raw x",
    x = NULL, 
    y = NULL
  ) 

```


## In Burke and Emerick

```{r BE_ex, fig.asp=0.8}
BE_data <- haven::read_dta(here(
  "misc", "burke_emerick_2016", "output", "panel_boot1.dta"))

BE_x <- BE_data |>
  transmute(
    y = logcornyield,
    x = higher,
    lower,
    prec_lo,
    prec_hi,
    fe = fips,
    year
  ) |> 
  remove_missing()

# compare FE and non FE
reg_no_fe <- feols(BE_x,  y ~ x, cluster = "fe")
reg_fe <- feols(data = BE_x, y ~ x | fe)

reg_no_fe
reg_fe
# with fe, larger standard errors (on each coef)

#why? Check that FE more correlated with x than y
# => here that FE remove more of the variance in (x, in proportion)
y_p_xfe <- residuals(reg_fe)
y_p_x <- residuals(reg_no_fe)
var(y_p_xfe)/var(y_p_x)

x_p_fe <- feols(BE_x,  x ~ 1 | fe) |> residuals()
x <- feols(BE_x,  x ~ 1, cluster = "fe") |> residuals()
var(x_p_fe)/var(x)

#plots
data_p <- tibble(y_p_xfe, y_p_x, x_p_fe, x, fips = BE_x$fe, year = BE_x$year) 

data_p |> 
  pivot_longer(cols = everything()) |> 
  filter(str_starts(name, "x")) |> 
  filter(name == "x") |> 
  mutate(name = ifelse(name == "x", " No Fixed Effects", "Fixed Effects")) |> 
  ggplot(aes(x = value, fill = name, color = name)) +
  # geom_density()
  geom_histogram(position = "identity", alpha = 0.3, fill = "#1C6C6F", color = "#1C6C6F") +
  labs(
    title = "The independent variable side in Burke and Emerick (2016)",
    subtitle = "",
    fill = NULL, 
    color = NULL,
    x = "Number of GDD above threshold",
    y = "Count"
  ) +
  xlim(c(-150, 500)) +
  ylim(c(0, 34000))

data_p |> 
  pivot_longer(cols = everything()) |> 
  filter(str_starts(name, "x")) |> 
  mutate(name = ifelse(name == "x", " No Fixed Effects", "Fixed Effects")) |> 
  ggplot(aes(x = value, fill = name, color = name)) +
  # geom_density()
  geom_histogram(position = "identity", alpha = 0.3) +
  labs(
    title = "The independent variable side in Burke and Emerick (2016)",
    subtitle = "Impact of county fixed effect",
    fill = NULL, 
    color = NULL,
    x = "Number of GDD above threshold",
    y = "Count"
  ) +
  xlim(c(-150, 500)) +
  ylim(c(0, 34000))

data_p |> 
  pivot_longer(cols = everything()) |> 
  filter(str_starts(name, "y_")) |> 
  mutate(name = ifelse(name == "y_p_x", " No Fixed Effects", "Fixed Effects")) |>
  ggplot(aes(x = value, fill = name, color = name)) +
  # geom_density()
  geom_histogram(position = "identity", alpha = 0.3) +
  labs(
    title = "The dependent variable side in Burke and Emerick (2016)",
    subtitle = "Impact of county fixed effect",
    fill = NULL, 
    color = NULL,
    x = "Log corn yield",
    y = "Count"
  )

```


Maps

Shape files

```{r}
library(maps)


#shape file
shape_counties <- map_data("county") |> 
  as_tibble() 

#need to map to county
data(county.fips)

county_fips_tidy <- county.fips |> 
  as_tibble() |> 
  separate(polyname, into = c("region", "subregion"), sep = ",") |> 
  mutate(fips = as.numeric(fips))

shape_fips <- shape_counties |> 
  full_join(county_fips_tidy, by = c("region", "subregion")) 
```


```{r}
normalize <- function(x) {
  (x - min(x, na.rm = TRUE))/(max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
} 

data_weights_BE <- data_p |> 
  filter(year == 1990) |> 
  # ungroup() |> 
  mutate(
    prop_x = abs((x_p_fe)/x),
    prop_y = abs((y_p_xfe)/y_p_x),
    x_norm = normalize(x),
    x_p_fe_norm = normalize(x_p_fe),
    weight_x = x_p_fe^2,
    weight_x_norm = weight_x/sum(weight_x, na.rm = TRUE),
    # weight_x_norm = normalize(weight_x_norm),
    weight_y = y_p_xfe^2,
    weight_y_norm = weight_y/sum(weight_y, na.rm = TRUE),
    prop_x_norm = x_norm/x_p_fe_norm,
    prop_y_p_x_norm = y_p_x_norm/y_p_xfe_norm
  ) |> 
  # group_by(year) |> 
  full_join(shape_fips, by = c("fips")) |> 
  filter(fips %in% unique(BE_data$fips))

data_weights_BE |> 
  ggplot() +
  geom_polygon(
    aes(x = long, y = lat, group = group, fill = weight_x_norm),
    colour = "white",
    size = 0.005
  ) + 
  coord_fixed(1.3) +
  labs(
    x = NULL, y = NULL
  ) +  
  theme_mediocre_map(pal = "coty") +
  theme(legend.position = "right") +
  scale_fill_continuous(guide = guide_colourbar(direction = "vertical")) +
  scale_mediocre_c(pal = "coty", gradient = "left") +
  labs(title = paste(unique(data_weights_BE$year)), sep = ", ")

data_weights_BE |> 
  pivot_longer(cols = c(x, x_p_fe)) |> 
  ggplot() +
  geom_histogram(aes(x = value, fill = name), position = "identity", alpha = 0.5)
```










