\documentclass[usletter, 12pt]{article}
 
\usepackage{VB}
\usepackage{cleveref}
\usepackage{amsthm}
\usepackage{mdframed}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}


%\doublespacing

\begin{document}

%\lhead{\textsc{Virtual paper - Public economics and development - Vincent Bagilet}}

\selectlanguage{english}
	
	\title{Causal Exaggeration}

	\author{Vincent Bagilet
		\thanks{Columbia University, New York, USA. Email: \url{vincent.bagilet@columbia.edu}. 
			A previous version of this project, entitled ``\textit{Unconfounded but inflated causal estimates}'', was co-authored with Léo Zabrocki-Hallak, I cannot thank him enough for his invaluable and far-reaching contributions to the project. I am very grateful to Hélène Ollivier and Jeffrey Shrader for their guidance and thank Sylvain Chabé-Ferret, Jesse McDevitt-Irwin, David McKenzie, José Luis Montiel Olea, Suresh Naidu, Claire Palandri, Julian Reif, Stephan Thies and Roberto Zuniga Valladares for helpful comments, as well as lab members at Columbia and seminars participants at Columbia, IPWSD, the Paris School of Economics and the Toulouse School of Economics.}
	}
	
	\institution{Sustainable Development PhD Program, Columbia University}

	\date{January 19, 2023}
	
	
	\maketitle
	
	\begin{center}
		\large \textsc{\textbf{Abstract}}\\
	\end{center}
		
		%To avoid confounding, causal identification strategies focus on a subset of the variation in the data, the plausibly exogenous part. In this paper, I argue that it can reduce statistical power and lead published estimates to exaggerate true effects sizes. Using realistic fake data simulations and a mathematical derivation, I show that for the main causal strategies, a perfectly convincing identification does not guaranty an absence of ``bias'' and that improving identification can actually pull us away from the true effect. I then discuss potential avenues to address this issue.\\~\\
		
		%Julian Reif's proposal, modified:
		The credibility revolution in economics has made causal inference methods ubiquitous. At the same time, an increasing amount of evidence has highlighted that the literature strongly favors statistically significant results. I show that these two phenomena interact in a way that significantly worsens the reliability of published estimates: while causal inference methods can successfully reduce bias caused by confounders, the accompanying reduction in statistical power can create another type of bias, leading published estimates to exaggerate true effects sizes. I document this trade-off using realistic fake data simulations and a formal mathematical derivation. I then discuss potential avenues to address it. 
		
		%Quasi-experimental studies make empirical economics credible. To avoid confounding, causal identification strategies focus on a subset of the variation in the data, the plausibly exogenous part. In this paper, I argue that it can reduce statistical power and lead published estimates to exaggerate true effects sizes. Using realistic fake data simulations and a mathematical derivation, I show that %for the main causal strategies, there can be a trade-off between confounding and exaggeration: we may avoid a bias at the risk of generating another one.
		%using the main causal identification strategies to avoid a bias can also generate another one. I then discuss potential avenues to address this issue.\\~\\
	
	
	
	\newpage
	
%-----------------------------------------------------------------------------

% INTRODUCTION

%-----------------------------------------------------------------------------
	
	\section{Introduction}
	
		\input{../preliminary_intro/causal_exaggeration_intro_text.tex}
	
	
%-----------------------------------------------------------------------------

% SIMULATIONS

%-----------------------------------------------------------------------------	
				
	\section{Simulations} \label{simulations}
	
%-----------------------------------------------------------------------------

% MATHS

%-----------------------------------------------------------------------------
	
	\newpage
	
	\section{Mathematical derivation} \label{maths}
						
		In this section, I formally prove the existence of the confounding / exaggeration trade-off  and describe its drivers. To do so, I first recall a general definition of the exaggeration ratio and discuss its evolution with the values of estimation parameters. Then, under a set of simplifying assumptions, I study the drivers of exaggeration for a series of estimators, in the presence of an omitted variable bias, when controlling for the omitted variable, when instrumenting it or running a reduced form analysis. Finally, I show that, for any magnitude of OVB, the overall bias combining exaggeration and OVB, can be greater when controlling for confounders using a causal inference methods than not. A more detailed analysis and proof is available on the \href{https://vincentbagilet.github.io/causal_exaggeration/Maths/math_causal_exaggeration.pdf}{companion website}.
		
%%%%%%%%%%%%%%  Exagg ratio  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		
		\subsection{Exaggeration ratio and its properties}\label{general_exagg}
		
			Following \cite{gelman_beyond_2014}, we can define the exaggeration ratio $E$ as follows:
		 
		 	\begin{definition}
		 		For an estimator $\hat{\beta}$ of a true effect of magnitude $\beta$, with standard deviation $\sigma$ an a two-sided test of size $\alpha$ with threshold value $z_{\alpha}$,
		 
				 \begin{equation}\label{exagg_general}
				 	E(\hat{\beta}, \sigma, \beta, z_{\alpha}) =
					 	\dfrac{\mathbb{E}\left[ |\hat{\beta}| \big| \beta, \sigma, |\hat{\beta}| > z_{\alpha} \sigma \right]}{|\beta|} 
				\end{equation}
			\end{definition}
			
			We can easily extend proofs by \cite{lu_note_2019} and \cite{zwet_significance_2021} to biased estimators, to show that, for a given size of a test and true effect size, exaggeration increases with the variance of a (un)biased estimator normally distributed estimator:
			
			\begin{lemma}\label{lemma_evol_exagg}
			
				For an estimator $\hat{\beta_{b}}  \sim \mathcal{N}(\beta + b, \sigma^{2})$ of a true effect of magnitude $\beta$ and a fixed bias $b$ of the same sign as and independent from the true effect,
				
				\begin{itemize}
					\item $E$ is a decreasing function of the Signal-to-Noise Ratio (SNR), $\frac{\beta}{\sigma}$, and only depends on $\sigma$ through this SNR. 
					\item  $\lim_{\sigma\to \infty} E(\hat{\beta_{b}}, \sigma, \beta, z_{\alpha}) = +\infty$.
				\end{itemize}
			\end{lemma}
			
				 We focus on cases in which the bias is in the same direction as the true effect to be conservative when comparing exaggeration from causal inference methods to exaggeration in the naive regression in the presence of OVB.\\
				 
				 Since the exaggeration ratio increases with the variance of the estimator, to study how it evolves with the IV strength in an IV setting, the number of exogenous shocks in a reduced form and the correlation between the explanatory variable of interest and the omitted variable of interest, we only have to study how the variance of these estimators evolves with these parameters.
				 				 
		\subsection{Setting and data generating process}\label{maths_dgp}
		
			Consider an usual linear homoskedastic regression model with an omitted variable. For all individual $i \in \{1, ..., n\}$, assume that: 
				~
				\begin{equation}\label{maths_true_dgp}
					y_i = \beta_{0} + \beta_{1}x_{i} + \delta w_{i} + u_i
				\end{equation}
				
				where $y_i \in \mathbb{R}$ is the observed outcome of interest, $x_i \in \mathbb{R}$ the observed explanatory variable of interest,  $w_i \in \mathbb{R}$ an unobserved omitted variable, $u_i \in \mathbb{R}$ an unobserved error term.  $(\beta_0, \beta_1, \delta) \in \mathbb{R}^{3}$ are unknown parameters. $\beta_1$ is the parameter of interest. For readability, we introduce the usual vector notation such that for instance $y = (y_1, ..., y_n)'$ and set $\bm{\beta} = (\beta_0, \beta_1)'$ and $\text{x}_i = (1, x_i)'$. \\
			
			We assume homogeneous treatment effects and homoskedasticity, along with the usual OLS assumptions (\textit{i.i.d.} observations, finite second moments, positive-definiteness of $\mathbb{E}[\text{x}_i \text{x}_i']$ and $u_{i}$ conditional mean-zero and uncorrelated with $x_i$ and $w_i$). We assume that $w_{i}$ is unobserved, correlated with $x_{i}$ and that $\delta \neq 0$. To simplify the derivations, we further assume that the unobserved variable is centred, \textit{i.e.} $\mathbb{E}[w_i] = 0$. We also assume that the variance of the component of $w_{i}$ that is orthogonal to $x_{i}$ (denoted $w_{i}^{\perp x}$) does not vary with $x_{i}$, \textit{i.e.}, $\text{Var}(w_{i}^{\perp x}|x_{i}) = \text{Var}(w_{i}^{\perp x})$. We assume the following data generating process for $x_i$:
				~
				\begin{equation}\label{maths_dgp_x}
					x_i = \mu_{x} + \gamma w_{i} + \epsilon_i
				\end{equation}
				
				where $\gamma \in \mathbb{R}^{*}$ since $x$ and $w$ are correlated such that $\rho_{xw} = \text{corr}(x, w) =  \frac{\gamma \sigma_{w}}{\sigma_{x}}$. Then, in the IV and reduced form sections, we further assume that there exists a valid instrumental variable $z_i$ for $x_i$, \textit{i.e.} that $\mu_x + \epsilon_{i} = \pi_0 + \pi_1 z_i + \epsilon_{\textsc{iv}, i}$  where $(\pi_0, \pi_1) \in \mathbb{R}^{2}$ are unknown parameters. The existence or not of this valid instrument does not affect the results in the controled and \textsc{OVB} cases. Since the instrument is valid, it satisfies exogeneity, \textit{ie} $\mathbb{E}[\text{z}_{i}u_{i}] = 0$ and  $\mathbb{E}[z_{i}w_{i}] = 0$, relevance, \textit{ie} $\text{rank}(\mathbb{E}[\text{z}_{i}\text{x}_i']) = 2$,  and positive-definiteness of $\mathbb{E}[\text{z}_i \text{z}_i']$. The data generating process for $x_i$ becomes:			
				~
				\begin{equation}\label{maths_dgp_x_iv}
					x_i = \pi_0 + \pi_1 z_i + \gamma w_{i} + \epsilon_{\textsc{iv}, i}
				\end{equation}
				
				We assume that $\epsilon_{\textsc{iv}, i}$ is uncorrelated with $z_i$ and $w_{i}$, \textit{ie} $\mathbb{E}[z_i\epsilon_{\textsc{iv}, i}] = 0$ and $\mathbb{E}[w_i\epsilon_{\textsc{iv}, i}] = 0$. We also assume homoskedasticity for this term, such that $\mathbb{E}[\epsilon_{\textsc{iv}, i}^{2} | z_{i}, w_{i}] = \sigma_{\epsilon_{\textsc{iv}}}^{2}$ is constant.\\ %It implies that  $\mathbb{E}[\epsilon_{\textsc{iv}, i}^{2} | z_{i}] = \sigma_{\epsilon_{\textsc{iv}}}^{2}$ (since $\mathbb{E}[\epsilon_{\textsc{iv}, i}^{2} | z_{i}] =  \mathbb{E}[\mathbb{E}[\epsilon_{\textsc{iv}, i}^{2} | z_{i}, w_{i}] | z_{i}] = \mathbb{E}[\sigma_{\epsilon_{\textsc{iv}}}^{2} | z_{i}] = \sigma_{\epsilon_{\textsc{iv}}}^{2})$.\\
				
				Overall, this DGP is close to the usual textbook one but with an additional omitted variable.

				
%%%%%%%%%%%%  ASYMPTOTIC DISTRIB  %%%%%%%%%%%%%%%%%%%%%%%%
		
		\subsection{Asymptotic distributions of the estimators}\label{maths_asymptotics}			 
			
			I now derive the asymptotic distributions of the various estimators and study the evolution of their variances with the value of the parameters of interest, \textit{i.e.} with a measure of the correlation between $x$ and $w$ ($\gamma$) in the controled / OVB case, of the IV strength ($\pi_{1}$) and of the number of exogenous shocks ($\sigma_{z}^{2}$ when $z$ is a dummy). %Since variables and parameters are intertwined, modifying the value of one of them can affect the value of the others. 
			In order for the variation of one factor not to impact other factors of interest, we fix the variances of variables ($\sigma_{y}^{2}, \sigma_{x}^{2}, \sigma_{w}^{2} \text{ and } \sigma_{z}^{2}$) but adjust for the variances of the error terms ($\sigma_{u}^{2} \text{ and } \sigma_{\epsilon}^{2}$) when varying the values of one of the parameters ($\gamma, \delta \text{ and } \pi_{1}$). This way, we basically vary how much the variance of $x$ and $y$ depends on ``defined'' variables ($y, x, w \text{ and } z$). Finally, to make things comparable across cases, I consider a constant bias, \textit{i.e.}, as shown below, $\gamma \delta = \kappa = cst$.
			
			
%%%%%%%%%%%%%%  OVB  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%

			\subsubsection{Naive regression (OVB)}\label{formal_proof_ovb}
			
				I first consider a ``naive'' regression of $y$ on $x$ (with $w$ omitted). Since, $w$ is unobserved, we consider the projection of $y$ on $X$ only, where $X = (\text{x}_{1}', ..., \text{x}_{n}')'$ with $\forall i, \text{x}_{i}' = (1, x_{i})$:
					~
					\begin{equation}\label{maths_eq_ovb}
						y = X\bm{\beta}_{\textsc{ovb}} + u_{\textsc{ovb}}
					\end{equation}
				
				with, by definition of the projection, $\mathbb{E}[X'u_{\textsc{ovb}}] = 0$.
						
				\begin{lemma}\label{lemma_ovb}
					Based on the data generating process described in section \ref{maths_dgp}, for $\hat{\beta}_{\textsc{ovb}}$ the OLS estimate of $\beta_{1}$ in the regression of $y$ on $x$, $\hat{\beta}_{\textsc{ovb}} \overset{d}{\to} \mathcal{N}(\beta_{1} + b_{\textsc{ovb}} , \ \sigma_{\textsc{ovb}}^{2})$, with
					\[
						b_{\textsc{ovb}} = \dfrac{\delta \gamma \sigma_{w}^{2}}{\sigma_{x}^{2}} 
						\qquad \text{and} \qquad
						\sigma_{\textsc{ovb}}^{2} = \dfrac{\sigma_{u}^{2} + \delta^{2} \sigma_{w}^{2}(1 - \rho_{xw}^{2})}{n \ \sigma_{x}^{2}}
						%\sqrt{n}\left( \hat{\beta}_{\textsc{ovb}} - \left(\beta_{1} + \dfrac{\delta \gamma \sigma_{w}^{2}}{\sigma_{x}^{2}}\right) \right)
						 %\overset{d}{\to} \mathcal{N}\left( 0 , \ \dfrac{ \sigma_{u}^{2} +  \delta^{2} \sigma_{w}^{2} (1 - \rho_{xw}^{2} )}{\sigma_{x}^{2}} \right) 
					\]
				\end{lemma}
				
			Note that $\sigma_{u}^{2} + \delta^{2} \sigma_{w}(1 - \rho_{xw}^{2})$ is the part of the variance in $y$ that is not explained by $x$ ($\sigma^{2}_{y^{\perp x}}$). We thus retrieved the variance described in introduction. Also note that since $\sigma_{x}^{2}$ and $\sigma_{w}^{2}$ are fixed, reasoning at $b_{\textsc{ovb}} = cst$ is equivalent to considering that $\gamma \delta = \kappa = cst$. Then, noting that $\forall i, u_{i} = y_i - \beta_{0} - \beta_{1}x_{i} - \delta w_{i}$ and computing its variance, we can then rewrite the $\sigma_{\textsc{ovb}}^{2}$ as a function of fixed variances and one or less varying parameter:
			~
			\[
				\sigma_{\textsc{ovb}}^{2} = \dfrac{\sigma_{y}^{2} - \beta_{1}^{2}\sigma_{x}^{2} - 2\beta_{1}\kappa\sigma_{w}^{2} - \kappa^{2}\frac{\sigma_{w}^{4}}{\sigma_{x}^{2}}}{n \ \sigma_{x}^{2}}
			\]
			
			 This expression underlines that, for a given bias,  $\sigma_{\textsc{ovb}}^{2}$ does not vary with $\gamma$, the parameter of interest. Applying lemma \ref{lemma_evol_exagg} proves that $E_{\textsc{ovb}}$ does not either.

%%%%%%%%%%%%%%  CTRL  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
			
			\subsubsection{Controlled regression}
			
				We then turn to the ``ideal'' case in which no variable is omitted, \textit{i.e.} we control for the omitted variable $w$. The model considered accurately represents the DGP: 
				~
				\begin{equation}\label{maths_eq_ovb}
					y = X_{w}\bm{\beta}_{\textsc{ctrl}} + u_{\textsc{ctrl}}
				\end{equation}
				
				where $X_{w} = (\text{x}_{w, 1}', ..., \text{x}_{w, n}')'$ with $\forall i, \text{x}_{w, i}' = (1, x_{i}, w_{i})$ and $\mathbb{E}[X_{w}'u] = 0$. This corresponds to the usual OLS setting with a constant and two regressors uncorrelated with the error.
						
				\begin{lemma}\label{lemma_ctrl}
					Based on the sus-mentioned data generating process, for $\hat{\beta}_{\textsc{ctrl}}$ the OLS estimator of $\beta_{1}$ in the regression of $y$ on $x$ and $w$, $\hat{\beta}_{\textsc{ctrl}} \overset{d}{\to} \mathcal{N}(\beta_{1}, \ \sigma_{\textsc{ctrl}}^{2})$, with
					\[
						\sigma_{\textsc{ctrl}}^{2} = \dfrac{\sigma_{u}^{2}}{n \ \sigma_{x}^{2} (1 - \rho_{xw}^{2})}
					\]
				\end{lemma}
				
				Note that $\sigma_{x}^{2} (1 - \rho_{xw}^{2})$ is the part of the variance of $x$ that is not explained by $w$ ($\sigma^{2}_{x^{\perp w}}$) and $\sigma_{u}^{2}$ the part of the variance of $y$ that is not explained by $x$ nor $w$ ($\sigma^{2}_{y^{\perp x, w}}$); here too we retrieved a result described in introduction. For a given bias, we then rewrite $\sigma_{\textsc{ctrl}}^{2}$ as a function of fixed variances and one varying parameter, $\gamma$:
				~
				\[
					\sigma_{\textsc{ctrl}}^{2} = \dfrac{\sigma_{y}^{2} - \beta_{1}^{2}\sigma_{x}^{2} - \frac{\kappa^{2}}{\gamma^{2}}\sigma_{w}^{2} - 2\beta_{1}\kappa \sigma_{w}^{2}}{n \ (\sigma_{x}^{2}  - \gamma^{2} \sigma_{w}^{2})}
				\]
				
				Since the numerator increases and the denominator decrease with $|\gamma|$,  $\sigma_{\textsc{ctrl}}^{2}$ increases with $|\gamma|$. For a given bias, the more $w$ is correlated with $x$ (and thus roughthly the less it is with $y$ since $\delta \gamma = cst$), the larger the variance of the estimator. In addition, we can note that, for a given bias, $\lim_{\gamma^{2} \to \frac{\sigma_{x}^{2}}{\sigma_{w}^{2}}} \sigma_{\textsc{ctrl}}^{2} = + \infty$.
				
			
%%%%%%%%%%%%%%  IV   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%	
		
		\subsubsection{Instrumental Variable}
		
			We now want to estimate the IV model in which we regress $y$ on $x$ instrumented by $z$. We instrument $\text{x}_{i} = (1, x_{i})'$ with $\text{z}_{i} = (1, z_{i})'$ and thus consider a just-identified case such that $\hat{\bm{\beta}}_{\textsc{2sls}} = \hat{\bm{\beta}}_{\textsc{iv}}$.						~
			\begin{lemma}\label{lemma_iv}
				Based on the sus-mentioned data generating process, for $\hat{\beta}_{\textsc{iv}}$ the IV estimate of $\beta_{1}$ in the regression of $y$ on $x$ instrumented by $z$,  $\hat{\beta}_{\textsc{iv}} \overset{d}{\to} \mathcal{N}(\beta_{1}, \ \sigma_{\textsc{iv}}^{2})$, with
					\[
						\sigma_{\textsc{iv}}^{2} = \dfrac{\sigma_{u}^{2} + \delta^{2}\sigma_{w}^{2}}{n \ \sigma_{x}^{2} \rho_{xz}^{2}}
					\]
			\end{lemma}
		
			For a given bias, noting that $\rho_{xz} = \text{corr}(x, z) = \pi_1 \frac{\sigma_z}{\sigma_x}$ and replacing $\sigma_{u}^{2}$, we can rewrite $\sigma_{\textsc{iv}}^{2}$ as a function of fixed variances and one varying parameter, $\pi_{1}$:
			\[
				\sigma_{\textsc{iv}}^{2} = \dfrac{\sigma_{y}^{2} - \beta_{1}^{2}\sigma_{x}^{2} - 2\beta_{1}\kappa \sigma_{w}^{2}}{n \ \pi_{1}^{2} \sigma_{z}^{2}}
			\]
			
			Clearly, the smaller $\pi_{1}$, the larger $\sigma_{\textsc{iv}}^{2}$. In addition, $\lim_{\pi_{1}\to 0} \sigma_{\textsc{iv}}^{2} = + \infty$.
			

			%We retrieve the usual result regarding the determinant of the 2SLS variance as discussed in \cite{hansen_econometrics_2022} for instance. %p354
			 %The variance of the estimator increases when $\sigma_u^{2}$, the variance of the error, increases. It decreases with  $\rho_{xz}$, the correlation between $x$ and $z$, and with $\sigma_{x}^{2}$ the variance of $x$, assuming the omitted variable fixed. 
				 
%%%%%%%%%%%%%%  Reduced form. %%%%%%%%%%%%%%%%%%	

		\subsubsection{Reduced form}
		
			Let's now assume that we want to directly estimate the effect of the instrument on the outcome of interest.
					While the DGP is analogous to that of the IV, since we directly regress the outcome on the instrument, the resulting estimand is different. To make them comparable, for the reduced form case, in equation \ref{maths_dgp_x_iv}, I set $\pi_{1}$ to 1. An increase of 1 in the instrument causes an increase of $\beta_{1}$ in $y$ and the estimands become comparable. All other assumptions and features of the DGP remain unchanged as compared to the IV. Under these assumptions, plugging equation \ref{maths_dgp_x_iv} into equation \ref{maths_true_dgp} yields:			
				\[
					y_{i} = \underbrace{\beta_{0} + \beta_{1}\pi_0}_{\beta_{0, \textsc{red}}} + \beta_{1}\pi_{1}z_{i} + \underbrace{(\delta + \beta_{1}\gamma) w_{i} + u_{i} + \beta_{1}\epsilon_{\textsc{iv}, i}}_{u_{\textsc{red}, i}}
				\]
				
				We thus estimate the following model:
				~
				\begin{equation}
					y_{i} =  \text{z}_{i}'\bm{\beta_{\textsc{red}}} + u_{\textsc{red}, i}
				\end{equation}
				
				where by the exogeneity assumption $\mathbb{E}[\text{z}_{i}u_{\textsc{red}, i}] = 0$.
				
				\begin{lemma}\label{lemma_red}
					Based on the sus-mentioned data generating process, for $\hat{\beta}_{\textsc{red}}$ OLS estimate of the reduced form regression of $y$ on $z$,  $\hat{\beta}_{\textsc{red}} \overset{d}{\to} \mathcal{N}(\beta_{1}, \ \sigma_{\textsc{red}}^{2})$, with
					\[
						\sigma_{\textsc{red}}^{2} = \dfrac{\sigma_{y}^{2} - \beta_{1}^{2}\pi_{1}^{2}\sigma_{z}^{2}}{n \ \sigma_{z}^{2}}
					\]
				\end{lemma}
				
				Note that as expected, the numerator is the part of the variance of $y$ that is not explained by $z$ ($\sigma^{2}_{y^{\perp z}}$). In addition, it is clear that the smaller $\sigma_{z}^{2}$, the larger $\sigma_{\textsc{red}}^{2}$. In addition, $\lim_{\sigma_{z}\to 0} \sigma_{\textsc{red}}^{2} = + \infty$.\\
				
				In the binary case, $\sigma_{z}^{2} = p_{1}(1-p_{1})$ with $p_{1}$ the proportion of treated observations, \textit{i.e.}, the proportion of 1 in $z$. When most observations have the same treatment status, \textit{i.e.}, $p_{1}$ close to 0 or 1, $\sigma_{z}^{2}$ tends to zero and $\sigma_{\textsc{red}}^{2}$ shoots up. There is not enough variation in the treatment status to precisely identify the effect of interest. 
				
		\subsection{Exaggeration ratios}
		
			Combining the results from lemma \ref{lemma_ovb} through \ref{lemma_red}  regarding the asymptotic distribution of the various estimators with lemma \ref{lemma_evol_exagg} stating the that exaggeration increases with the variance of a normally distributed estimator, we have:
			~
			\begin{theorem}
				For the data generating process described in section \ref{maths_dgp}, the exaggeration ratio of the controled, IV and reduced form estimators, respectively $E_{\textsc{ctrl}}, E_{\textsc{iv}} \text{ and } E_{\textsc{red}}$, are such that:
					\begin{itemize}
						\item $E_{\textsc{ctrl}}$ increases with the correlation between the omitted variable and the explanatory variable of interest (\textit{i.e.} $|\gamma|$ or $|\rho_{xw}|$), for a given bias,
						\item $E_{\textsc{iv}}$ increases with the strength of the IV (\textit{i.e.} with $|\pi_{1}|$ or $|\rho_{xz}|$),
						\item $E_{\textsc{red}}$ increases when the number of exogenous shocks decreases in the binary case
					\end{itemize}
			\end{theorem}
			
			Also using the same lemma and the limit properties of the variances described in section \ref{maths_dgp}, and since, at fixed bias, $E_{\textsc{ovb}}$ does not vary with the parameters of interest, we have:
			~
			\begin{theorem}
				For the data generating process described in section \ref{maths_dgp}, 
					\begin{itemize}
						\item $\forall \ b_{\textsc{ovb}}, \exists \gamma$ s.t. $E_{\textsc{ctrl}} > E_{\textsc{ovb}}$
						\item $\exists \ \pi_{1}$ s.t. $E_{\textsc{iv}} > E_{\textsc{ovb}}$
						\item $\exists \ \sigma_{z}$ s.t. $E_{\textsc{red}} > E_{\textsc{ovb}}$
					\end{itemize}
			\end{theorem}
			
			For some parameter values, statistically significant estimates can be, on average larger when using a convincing causal identification strategy that eliminates the omitted variable bias than when embracing the bias and running a naive biased regression.
						
%-----------------------------------------------------------------------------

% PRACTICAL RECOMMENDATIONS

%-----------------------------------------------------------------------------

	\section{Practical Recommendations} \label{discussion}

		% HOOK OF THE SECTION
 		We have seen that using causal methods induces a trade-off between avoiding confoundings and exaggerating true effects if the study is under-powered. How can we, as applied researchers using observational data, arbitrate it? Since key pieces of information such as the true effect and the effect of omitted variables are inherently unknown, we cannot directly compute the biases caused by confounders and exaggeration. In this section, I examine how we can however get a sense of threats from both sides of the trade-off and visualize its main driver, the variation used for identification. I then discuss how changing attitudes towards statistical significance and replicating studies could limit the exaggeration issue.
	
	%Even though it does not produce uninflated estimates, reporting power calculations enables to evaluate the risk of exaggeration for a study. In this section, we present a workflow to evaluate and report the power of a study before and after its implementation. 
	
		\subsection{Gauging omitted variable bias}
	
			%"MEASURING" OVB
			On one side of the trade-off lies the bias caused by confounders. Although it is in essence impossible to measure, tools such as sensitivity analyses are available to gauge its magnitude \citep{rosenbaum_observational_2002, middleton_bias_2016, oster_unobservable_2019, cinelli_making_2020}. For instance, the method developed in \cite{cinelli_making_2020} enables to assess how strong confounders would have to be to change the estimate of the treatment effect beyond a given level we are interested in. It offers bounds for the strength of the association between the treatment and potential omitted variables by weighting it against the measured association between the treatment and observed covariates. A typical conclusion from such an analysis would be: ``omitted variables would have to explain as much residual variance of the outcome and the treatment as the observed covariate $x$ to bring down the estimate to a value of $\beta_{l}$''. In addition, the authors implement graphical tools to facilitate this comparison. I suggest to use such quantitative bias analyses to evaluate the restrictiveness of the causal approach required to limit the threat of unobserved confounding to acceptable levels. 
			
		\subsection{Evaluating risks of exaggeration}

			On the other side of the trade-off lies the exaggeration emerging when statistical power is low. As OVB, exaggeration and statistical power are in essence impossible to measure as their computation depends on the true effect which is always unknown. Yet, power calculations can help assess them by making hypothesizes on the magnitude of the true effect. In randomized controlled trials, such computations are not only an established practice but a requirement \citep{duflo_using_2007, mcconnell_going_2015, athey_econometrics_2016}. They are, however, rarely reported in non-experimental studies. This can be understandable if power calculations are perceived only as a way to measure a design's ability to detect an effect when there is actually one. Yet, taking publication bias and the threat of exaggeration into account highlights the necessity of running power calculations in non-experimental studies as well. A low power or a relatively large variance not only makes it more difficult to detect an effect or to draw clear conclusions about its magnitude when detected but it can also create a bias. Considering the attention given to bias avoidance in the economics literature, I advocate to make power central to non-experimental analyses to limit bias caused by exaggeration. Currently, in causal inference textbooks, very few pages are devoted to statistical power in non-experimental studies \citep{angrist_mostly_2009, angrist_mastering_2014, imbens_causal_2015, cunningham_causal_2021}. To the best of my knowledge, only two textbooks discuss the matter in depth \citep{shadish_experimental_2002, huntington-klein_effect_2021}. Results from power and exaggeration calculations would not only be highly informative but could also be reported very concisely in the robustness section of articles. 
			
			\subsubsection{Prospective power calculations}
						
				To evaluate the statistical power of a study, the risk of exaggeration and identifying the factors driving it, one can first simulate the design of the study \citep{hill_bayesian_2011, gelman_regression_2020, black_simulated_2021}. Simulating a data generating process from scratch requires thinking about the distribution of the variables, about their relationships and can also help underline the variation used for identification. %External information found in previous studies can help guide the simulation process to make it more realistic. 
			 I implemented such fake-data simulations in \Cref{simulations} and provide "R" code to be used as examples in the replication material. In situations where the relationships among covariates are too complex to emulate, one can also start from an existing dataset and add a known treatment effect. I implemented real data simulations in a companion paper and describe their implementation in its \href{https://vincentbagilet.github.io/inference_pollution/}{replication material} \citep{bagilet_accurately_2023}. 
			 
			 %When simulations indicate that statistical power is low, additional data could be collected or the statistical model could be improved to increase precision. In any case, it should not stop the research project from carrying out. Simulation results rest on the way the data generation process was modeled, and it can be difficult to gauge the amount of noise present in data before actually analyzing them. The two main benefits of a prospective simulation procedure are that it allows us to consider factors that may affect the statistical power of our study and to avoid drawing misleading conclusions based solely on statistically significant estimates when statistical power is low.
				
			\subsubsection{Retrospective power calculations}
			
				Running post-analysis power calculations can also help getting a sense of the statistical power associated with a research design. Such \textit{retrospective} calculations allow to evaluate whether the design of the study would produce accurate and uninflated statistically significant estimates if the true effect was in fact smaller than the observed estimate \citep{gelman_beyond_2014, ioannidis_power_2017, stommes_reliability_2021}. 
				
				%can take this out if want to make the paper shorter
				I illustrate how a retrospective analysis works by taking the example of \cite{card_using_1993} on the relationship between human capital and income. He finds that an additional year of education, instrumented by the distance of growing near a four-year college, causes a 13.2\% average increase in wage. The associated standard error is 5.5\%. Is there a risk of exaggeration with this design? Since, as noted by the author himself, the estimate is very imprecise we could expect so. Imagine that prior evidence suggests that the true effect is likely close to a 10\% increase in wage. We can then compute the statistical power of the study by drawing many estimates from a normal distribution centered around the hypothesized true effect of 10\% and with a standard deviation equal to the 5.5\% standard error obtained in \cite{card_using_1993}. Concretely, this proceeds as if we were able to replicate the study many times under the assumption that the true effect was equal to the hypothesized one\footnote{\cite{timm_retrodesign_2019, linden_retrodesign_2019} offer \texttt{R} and \texttt{Stata} packages that enable to easily run these calculations through a short command: \texttt{retrodesign(10, 5.5)}.}. Statistically significant estimates (at the 5\% level) would on average be roughly equal to 15\%, therefore overestimating the true effect by a factor of 1.5. Statistical power, the proportion of estimates that are significant, would be only 44\%. Conditional on a 10\% true effect size being a reasonable assumption, this study would be under-powered and the risk of exaggeration large.
			
				The usefulness of any retrospective power analysis lies on the assumption regarding the true effect size. Informed guesses about the range of plausible effect sizes can be based on results from meta-analyses or previous studies with a convincing design (\textit{e.g.}, a large randomized controlled trial)\footnote{Note that when such meta-analyses are available, one can use a Bayesian procedure to shrink statistically significant estimates based on the corpus of estimates from prior studies \cite{zwet_proposal_2021, zwet_significance_2021, zwet_statistical_2021} .}. When such information is not available, power calculations can be ran for a range of smaller but credible effect sizes.
				
		\subsection{Driver of the trade-off}
		
			We have seen that the variation used for identification was the main driver of the confounding / exaggeration trade-off. This variation determines the variance of the estimator and thus the amount of exaggeration. In this section, I first discuss the impact of estimator variance and how exaggeration leads to reinterpret the well-known bias/variance trade-off. I then propose to use a tool that enables to visualize where the variation used for identification comes from.
		
			\subsubsection{Navigating the bias/variance trade-off}
			
				In non-experimental studies, estimator variance is often important to the extent that a large variance may lead to a failure to reject the null of zero effect when it is actually incorrect. It makes variance paramount as long as a statistically significant estimate is not obtained. Yet, exaggeration underlines that variance matters, even after that a significant estimate has been obtained. 
			
				 Obtaining a statistically significant estimate from an imprecise estimator should not necessarily be interpreted as a sign of ``success'' in getting significance despite the estimator being imprecise. It could however be a warning that this estimate comes from the tails of the distribution and thus inaccurately represents the true effect. Conditional on having obtained a statistically significant estimate, a limited precision can hide a bias: exaggeration. This invites to revisit the well-known bias/variance trade-off: a larger variance can actually lead to a larger (conditional) bias. The bias/variance trade-off can in fact be a bias/bias trade-off. This paper thus invites to pay more attention to the implications of our design choices on the variance of our estimators, even after we obtained statistically significant estimates.
			
			\subsubsection{Visualizing the driver of the trade-off}
		
				Causal inference strategies only leverage a subset of the variation to avoid confounders. However, when this subset is too small, exaggeration arises. Visualizing this variation (and the observations) actually used for identification can help navigate the confounding / exaggeration trade-off. The visualization tool I outline here both leverages the interpretation of causal inference methods as control approaches and builds on a procedure developed in \cite{aronow_does_2016} for a different purpose: evaluating external validity of standard regressions.
				
				\cite{aronow_does_2016} essentially interprets the estimate of the coefficient of the treatment of interest in a simple linear non-causal regression as a weighted average of individual treatment effects. The weight $w_{i}$ of individual $i$ is simply the squared difference between its treatment status $T_{i}$ and the value of this treatment status as predicted by the other covariates $X$: $w_{i} = (T_{i} - \mathbb{E}[T_{i} | X_{i}])^{2}$. If treatment effects are heterogenous, the weighting may lead some observations to be disproportionally (non-)represented in the average effect of the treatment. %Some observations, whose treatment status is very well explained by covariates do not actually contribute to the estimation of the treatment effect. 
				In that case, the average of treatment is only representative of a subset of the individual treatments, leading to external validity issues. 
				
				The parallel with our setting directly follows from this interpretation regardless of whether the treatment is heterogenous or not: observations whose treatment status is well explained by covariates do not actually contribute to the estimation of the effect of the treatment. This may lead to a small \textit{effective} sample size and to exaggeration. In the control approach to causal inference strategies, the more variation in the treatment is absorbed when ``controlling'' for confoundings, the smaller the effective sample, potentially leading to exaggeration. \cite{aronow_does_2016} focused on the representativity of the effective sample for fear of external validity issues. I am interested on in its size for fear of exaggeration.
				
				It might then seem compelling to define this effective sample by proposing a weight value under which the associated observation does not actually contribute to  identification. Yet, considering the specificity of each analysis and that exaggeration depends on several factors, including the true effect size, I avoid proposing an \textit{ad hoc} cut-off. Instead, I suggest to visualize the individual weights. It allows to get a sense of where the variation comes from and which are the observations that actually contribute to the estimation. Since applied economics analyses often rely on panel data, I propose to use a heatmap as a base for visualization, with time on the \textit{x}-axis and individuals on the \textit{y}-axis. Figure %
				displays this visualization for the paper
				
				Most of the variation is absorbed by fixed effects, leading only a small subset of observation to actually contribute to the estimation (the darker ones).
				
				If the data is cross sectional, the same heatmap can be computed with individuals on the \textit{x}-axis and a length-one segment on the \textit{y}-axis. I display such graphs and share  \verb?R?  code to build on the \href{https://vincentbagilet.github.io/causal_exaggeration/}{companion website}.
				
				%talk about leverage
				
	% GENERAL RECOMMENDATIONS
	\subsection{Attitude Towards Statistical Significance and Replication}
	
		Exaggeration only arises in the presence of publication bias. As shown in the simulations, if estimates were not filtered by their statistical significance, even under-powered studies would on average recover the true effect, as long as the estimator is unbiased. The exaggeration issue could there be addressed by tackling publication bias. 
		
		To identify broader pathways to eliminate this filtering of significant results, it is first helpful to discuss the processes that lead to statistically significant results when power and thus the probability of obtaining a significant estimate is low. In such situations, they can be obtained either by ``chance'' or as an outcome of the garden of forking paths \citep{simmons_false-positive_2011, gelman_garden_2013, kasy_forking_2021}. Forks appear at various stages along the path of research, for instance in data preparation, regarding the inclusion of a given control variable or whether to carry on a research that yields non-significant results. Due to the structural flaw that favors significance, the path followed may be more likely to lead to a statistically significant result. These choices are most often not the result of bad researcher practices but instead a product of a structure that portrays significant results as the end goal of research. 
		
		The issue being structural, system level changes in scientific practices could also alleviate exaggeration and the trade-off described in this paper. First, many researchers advocate abandoning statistical significance as a measure of a study's quality \citep{mcshane_abandon_2019}. To be effective, this change should be paired with an effort to replicate studies \citep{christensen_transparency_2018}. Replications, even of low powered studies, would eventually enable to build the actual distribution of the causal estimand of interest. Meta-analyzes would then reduce the uncertainty around the true value of the causal estimand by pooling estimates \citep{hernan_causal_2021}. Finally, the inflation of statistically significant estimates could be limited by considering confidence intervals as compatibility intervals \citep{shadish_experimental_2002, amrhein_inferential_2019, romer_praise_2020}. The width of these intervals gives a range of effect sizes compatible with the data. Confidence intervals will be wide in under-powered studies signaling that point estimates should not be taken at face value, even if statistically significant.

%-----------------------------------------------------------------------------

% CONCLUSION

%-----------------------------------------------------------------------------

\section{Conclusion} \label{conclusion}

	The economic literature suffers from an extensive lack of statistical power \citep{ioannidis_power_2017} and many studies showed that this literature strongly favors statistically significant findings \citep[for instance]{rosenthal_file_1979, andrews_identification_2019, abadie_statistical_2020, brodeur_methods_2020}. Yet, in the presence of such a publication bias, estimates published from underpowered studies exaggerate true effect sizes, even when the estimators are ``unbiased'' in the usual sense of $\mathbb{E}[\hat{\beta}] = \beta$ \citep{ioannidis_why_2008, gelman_beyond_2014, lu_note_2019, zwet_significance_2021}. It is therefore not surprising that many estimates published in economics have been shown to be considerably exaggerated  \citep{camerer_evaluating_2016, ioannidis_power_2017}, despite the extensive use of convincing causal inference methods.  However, determinants for these exaggeration and power issues remain understudied. Focusing on causal methods, I argue that exaggeration may be exacerbated by the foundational component of causal inference: the fact that it only leverages subsets of the variation. Although it enables to avoid confounding, it also reduces statistical power and thus increases the risk of exaggeration. The same aspect that makes these methods credible can create another type of bias. A systematic reporting of statistical power calculations and analysis of the variation actually used for identification could help avoid falling into this exaggeration trap.
	
		
	
\newpage
	
\bibliographystyle{agsm}
\bibliography{../causal_exaggeration}

\newpage

\appendix

	\section{Mathematical proofs}
	
%%%%%%%%%% Evol exaggeration ratio %%%%%%%%%%%%%%%%

		\subsection{Variation of the exaggeration ratio (lemma \ref{lemma_evol_exagg})}
			
			\begin{proof}
				\cite{lu_note_2019} and \cite{zwet_significance_2021} showed this in the case of $b = 0$. 
				To extend it to the biased case, consider 
				$E_{b} = \frac{\mathbb{E}\left[ |\hat{\beta}_{b}| \big| \beta_{1}, \sigma, |\hat{\beta}_{b}| > z_{\alpha} \sigma \right]}{|\beta_{1}|}$ the exaggeration ratio of interest. 
				Note that, since $\hat{\beta}_{b}$ is an unbiased estimator of $\beta_{1} + b$, $\tilde{E_{b}} = \frac{\mathbb{E}\left[ |\hat{\beta}_{b}| \big| \beta_{1}, \sigma, |\hat{\beta}_{b}| > z_{\alpha} \sigma \right]}{|\beta_{1 + b}|}$ has the properties described in the lemma. 
				Now, considering that $E_{b} = \left| \frac{\beta_{1} + b}{\beta_{1}} \right| \tilde{E_{b}}$ proves the properties when $\beta_{1}$ and $b$ have the same sign.
			\end{proof}
	
%%%%%%%%%% Distrib CTRL  %%%%%%%%%%%%%%%%

\subsection{Asymptotic distribution of $\hat{\beta}_{\textsc{ctrl}}$ (lemma \ref{lemma_ctrl})}
	
	\begin{proof}
				The proof is the well know proof of the asymptotic distribution of the OLS. I simply compute $\mathbb{E}[x_{w, i}x_{w, i}']^{-1}$ to retrieve the variance of the parameter of interest $\bm{\beta}_{\textsc{ctrl}}$. We know that we have:
				~
				\[ \sqrt{n}(\hat{\bm{\beta}}_{\textsc{ctrl}} - \bm{\beta}_{\textsc{ctrl}}) \overset{d}{\to} 
					\mathcal{N}\left(0,  \mathbb{E}[\text{x}_{w, i}\text{x}_{w, i}']^{-1} \sigma_{u}^{2} \right) \]
			
			We are interested in the second component of $\hat{\bm{\beta}}_{\textsc{ctrl}}$. To retrieve it we need to compute $\mathbb{E}[\text{x}_{w, i}\text{x}_{w, i}']^{-1}$. 
			~
			\[ 
				\mathbb{E}[\text{x}_{w, i}\text{x}_{w, i}'] 
				= \mathbb{E} \begin{bmatrix}
					1 & x_{i} & w\\ 
					x_{i} & x_{i}^{2} & x_{i}w_{i}\\\
					w_{i} & x_{i}w_{i} & w_{i}^{2}
				\end{bmatrix} 
				=
%				\begin{bmatrix}
%					1 & \mu_{x} & 0\\ 
%					\mu_{x} & \sigma_{x}^{2} + \mu_{x}^{2} & \mathbb{E}[xw]\\
%					0 & \mathbb{E}[xw] & \sigma_{w}^{2}
%				\end{bmatrix} 
%				=
				\begin{bmatrix}
					1 & \mu_{x} & 0\\ 
					\mu_{x} & \sigma_{x}^{2} + \mu_{x}^{2} & \gamma\sigma_{w}^{2}\\
					0 & \gamma\sigma_{w}^{2} & \sigma_{w}^{2}
				\end{bmatrix} 
				\]
				
				Note that we have $\mathbb{E}[x_{i}w_{i}] = \mathbb{E}[x_{i}]\underbrace{\mathbb{E}[w_{i}]}_{0} + \text{cov}(x_{i}, w_{i}) = \gamma  \underbrace{\text{var}(w_{i})}_{\sigma_{w}^{2}} + \underbrace{\text{cov}(\epsilon_{i}, w_{i})}_{0} = \gamma\sigma_{w}^{2}$.\\
				
				 Now, $\mathbb{E}[\text{x}_{w, i}\text{x}_{w, i}']^{-1} = \dfrac{1}{\text{det}(\mathbb{E}[\text{x}_{w, i}\text{x}_{w, i}'])}~^{t}\text{C}$ with C the comatrix of $\mathbb{E}[\text{x}_{w, i}\text{x}_{w, i}']$. We have:
				~
				\[
					\text{det}(\mathbb{E}[\text{x}_{w, i}\text{x}_{w, i}']) = (\sigma_{x}^{2} + \mu_{x}^{2})\sigma_{w}^{2} -\sigma_{w}^{2}\mu_{x}^{2} - \gamma^{2}\sigma_{w}^{4} = \sigma_{w}^{2}(\sigma_{x}^{2} - \gamma^{2}\sigma_{w}^{2})
				\]
				
				and the ``central'' component of C, $\sigma_{w}^{2}$. Thus the central component of interest of $\mathbb{E}[\text{x}_{w, i}\text{x}_{w, i}']^{-1}$ is $\frac{1}{\sigma_{x}^{2} - \gamma^{2}\sigma_{w}^{2}}$. Therefore, for $\hat{\beta}_{\textsc{ctrl}}$, the second component of $\hat{\bm{\beta}}_{\textsc{ctrl}}$, we have:
				\begin{equation}
						\hat{\beta}_{\textsc{ctrl}} \overset{d}{\to}
							 \mathcal{N}\left( \beta_1 , \ \dfrac{\sigma_{u}^{2}}{n \ (\sigma_{x}^{2} - \gamma^{2}\sigma_{w}^{2})} \right) 
				\end{equation}
				
				Then, noting that $\text{cov}(x, w) = \text{cov}(\mu_{x} + \gamma w + \epsilon, w) = \gamma \sigma_{w}^{2}$ and thus $\text{corr}(x, w) = \frac{\text{cov}(x, w)}{\sigma_{x}\sigma_{w}} = \frac{\gamma\sigma_{w}}{\sigma_{x}} = \rho_{xw}$. We have:
				\[
					\sigma_{\textsc{ctrl}}^{2} = \dfrac{\sigma_{u}^{2}}{n \ \sigma_{x}^{2}(1 - \rho_{xw}^{2})}
				\]
				~
				\end{proof}
	
%%%%%%%%%% Distrib IV  %%%%%%%%%%%%%%%%

\subsection{Asymptotic distribution of $\hat{\beta}_{\textsc{iv}}$ (lemma \ref{lemma_iv})}
	
	\begin{proof}
			Since $u_{\textsc{iv}} = u_{\textsc{ovb}} = \delta w + u$, we have $\sigma_{u_{\textsc{iv}}}^2 = \sigma_{u}^{2} + \delta^{2}\sigma_{w}^{2}$. Thus, the usual asymptotic distribution of the IV gives:
				~
				\[ \sqrt{n}(\hat{\bm{\beta}}_{\textsc{iv}} - \bm{\beta}) \overset{d}{\to} 
					\mathcal{N}\left(0,  (\sigma_{u}^{2} + \delta^{2}\sigma_{w}^{2}) \mathbb{E}[\text{z}_{i}\text{x}_{i}']^{-1} \mathbb{E}[\text{z}_{i}\text{z}_{i}'] \left( \mathbb{E}[\text{z}_{i}\text{x}_{i}']^{-1}\right)' \right) \]
					
			We are interested in the second component of $\hat{\bm{\beta}}_{\textsc{iv}}$. To retrieve it we need to compute $\mathbb{E}[\text{z}_{i} \text{z}_{i}]$, $\mathbb{E}[\text{x}_{i}\text{z}_{i}']^{-1}$ and its transpose.  
			~
			\[ 
				\mathbb{E}[\text{z}_{i}\text{z}_i'] 
				= \mathbb{E} \begin{bmatrix}
					1 & z_{i}\\ 
					z_{i} & z_{i}^{2}
				\end{bmatrix} 
				=
				\begin{bmatrix}
					1 & \mu_{z}\\ 
					\mu_{z} & \sigma_{z}^{2} + \mu_{z}^2
				\end{bmatrix} 
				\]
			
			\[ 
				\mathbb{E}[\text{z}_{i}\text{x}_i'] 
				=
				\begin{bmatrix}
					1 & \mathbb{E}[x_{i}]\\ 
					\mathbb{E}[z_{i}] & \mathbb{E}[z_{i}x_{i}]
				\end{bmatrix} 
				=
				\begin{bmatrix}
					1 & \pi_0 + \pi_1 \mathbb{E}[z_{i}] + \gamma \underbrace{\mathbb{E}[w_{i}]}_{0} + \underbrace{\mathbb{E}[\epsilon_{\textsc{iv}, i}]}_{0}\\ 
					\mu_{z} & \pi_0 \mathbb{E}[z_{i}] + \pi_1 \mathbb{E}[z_{i}^{2}] + \gamma \underbrace{\mathbb{E}[z_{i}w_{i}]}_{0} + \underbrace{\mathbb{E}[z_{i}\epsilon_{\textsc{iv}, i}]}_{0}
				\end{bmatrix} =
				\begin{bmatrix}
					1 & \pi_0 + \pi_1 \mu_{z}\\ 
					\mu_{z} & \pi_0\mu_z + \pi_1 (\sigma_{z}^{2} + \mu_{z}^{2})
				\end{bmatrix} 
			\]
			\[
				 \Rightarrow \quad
				     \mathbb{E}[\text{z}_{i}\text{x}_i']^{-1} = \dfrac{1}{\pi_1 \sigma_{z}^{2}} 
				     			\begin{bmatrix}
								 \pi_0 \mu_z + \pi_1 (\sigma_{z}^{2} + \mu_{z}^{2})& - \pi_0 - \pi_1 \mu_{z}\\ 
								 -\mu_{z} & 1
							\end{bmatrix} 
			\]
			
		Thus, 
			\[
				     \mathbb{E}[\text{z}_{i}\text{x}_i']^{-1}\mathbb{E}[\text{z}_{i}\text{z}_i']  \left( \mathbb{E}[\text{z}_{i}\text{x}_i']^{-1} \right)' 
				     		= \dfrac{1}{\pi_1 \sigma_{z}^{2}} 
				     			\begin{bmatrix}
								2 \pi_0 \mu_z + \pi_1(\sigma_{z}^{2} + \mu_{z}^{2}) + \frac{\pi_0^2}{\pi_1}&   -\mu_{z} -  \frac{\pi_0}{\pi_1} \\ 
								  -\mu_{z} -  \frac{\pi_0}{\pi_1} & \frac{1}{\pi_1}
							\end{bmatrix} 
			\]
			
			And so, for $\hat{\beta}_{\textsc{iv}}$, the second component of $\hat{\bm{\beta}}_{\textsc{iv}}$, we have:
				\begin{equation}
					\sqrt{n}\left( \hat{\beta}_{\textsc{iv}} - \beta_{1} \right)
					\overset{d}{\to}
					\mathcal{N}\left( 0 , \ \dfrac{\sigma_{u}^{2} + \delta^{2}\sigma_{w}^{2}}{n \ \pi_{1}^{2}\sigma_{z}^{2}} \right) 
				\end{equation}
				~
				Now, since $ \rho_{xz} = \text{corr}(x_{i}, z_{i}) =  \frac{\text{cov}( \pi_0 + \pi_1 z_i + \gamma w_{i} + \epsilon_{\textsc{iv}, i} , z_i)}{\sigma_x \sigma_z} = \pi_1 \frac{\sigma_z}{\sigma_x}$,
				~
				\[
						\sqrt{n}\left( \hat{\beta}_{\textsc{iv}} - \beta_{1} \right)
						\overset{d}{\to}
							 \mathcal{N}\left( 0 , \ \dfrac{\sigma_{u}^{2} + \delta^{2}\sigma_{w}^{2}}{\sigma_{x}^{2} \rho_{xz}^{2}} \right) 
				\]
			\end{proof}
				
%%%%%%%%%% Distrib RED  %%%%%%%%%%%%%%%%


\subsection{Asymptotic distribution of $\hat{\beta}_{\textsc{red}}$ (lemma \ref{lemma_red})}
				
			\begin{proof}
				The proof is straightforward: this is the usual univariate, unbiased case, with and error term equal to $(\delta + \beta_{1}\gamma) w_{i} + u_{i} + \beta_{1}\epsilon_{\textsc{red}, i}$. Since $w$, $u$ and $\epsilon_{\textsc{red}}$ uncorrelated, its variance is $(\delta + \beta_{1}\gamma)^{2} \sigma_{w}^{2} + \sigma_{u}^{2} + \beta_{1}^{2}\sigma_{\epsilon_{\textsc{red}}}^{2}$.
			\end{proof}
		
\end{document}